---
title: "Exploring The Jail Populatiom"
author: "Sam Powers"
date: "1/9/2019"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(knitr)
library(RColorBrewer)
library(xlsx)
crime <- read_csv("colorado_crime_stats.csv")
population <- read_csv("colorado_population_2008-2017.csv")[,-1] # File details the ACS Population estimates for colorado population 2008-2017
incarceration <- read.xlsx("colorado-incarceration-hist.xlsx",1, header = FALSE)[,-3] # Details Jail population by year from the Vera incarceration-trends data set
co_vera_data <-  read_csv("Colorado-Incarceration-Data.csv") # Details all Vera incarceration-trends data set for state of Colorado
census_join <- read_csv("vera_census_join.csv")


```

# Crime Rates
## Look at various rates of crime as a proportion of the population 
This is colorado-wide data 


```{r}
crime.agg <- crime %>% group_by(year, class) %>% summarise(total = sum(total)) # This aggregates the number of crimes by class by year 

crime.per <- crime.agg %>% 
  inner_join(population, by = "year") %>% 
  mutate(crime.percent = total/pop_estimate *100 ) %>%
  select(year, class, crime.percent)

crime.rate.vis <- ggplot(crime.per, aes(x= year, y = crime.percent, color = class)) +
  geom_line() + 
  geom_point()+
  ggtitle('Crime Rate by Type Over Time') +
  theme(
    plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
    axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
    axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
    axis.text.x=element_text(angle=50, size=10, vjust=0.5, family ="serif"),
    axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
    panel.background = element_rect(fill = 'white'),
    panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
    panel.grid.minor = element_line(colour = "white", size = .5),
    legend.text = element_text(family = "serif"), 
    legend.title = element_text(family = "serif", face = "bold"),
    legend.position = c(0.9, 0.50))+ 
  ylim(0, 4.0) +
  scale_x_continuous(breaks = seq(2008, 2017, 1), limits = c(2008,2017)) +
  labs(x="Year", y="Crime Rate (% of State Population)" , color = "Type of Crime")
crime.rate.vis

```

# Incarceration Rates
## This shows the Overall Incarceration Rate in Colorado over the Past few years
Note that there is not a whole lot of difference between 2008 and 2017. 
Also note this is with the flawed data (years are wrong, maybe). 
Percentages are expressed as percent of population ages 15-64

```{r}
names(incarceration) <- c("year", "jail_pop" )


co_jail_pop_by_year <- co_vera_data %>% 
  select(year, total_jail_pop, female_jail_pop, male_jail_pop, black_jail_pop, white_jail_pop, total_pop_15to64, total_pop, total_jail_pretrial) %>%
  group_by(year)%>%
  summarise_all(funs(sum(., na.rm = TRUE)))

incarc.pop <- incarceration %>% 
  left_join(select(co_jail_pop_by_year, year, total_pop_15to64, total_pop), by = "year")
#incarc.pop[incarc.pop$year == 2017, c("total_pop_15to64","total_pop")] <- c(3742397,5530105)

incarceration.perc.co <- incarc.pop %>% 
    mutate(jail.perc = jail_pop/total_pop_15to64 *100 ) %>%
    select(year, jail.perc)

incarc.rate.vis <- ggplot(incarceration.perc.co, aes(x= year, y = jail.perc)) +
  geom_line() + 
  geom_point()+
  ggtitle('CO Incarceration Rate Over Time') +
  theme(
    plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
    axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
    axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
    axis.text.x=element_text(angle=50, size=10, vjust=0.5, family ="serif"),
    axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
    panel.background = element_rect(fill = 'white'),
    panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
    panel.grid.minor = element_line(colour = "white", size = .5)
  ) + 
  ylim(0, .5) +
  scale_x_continuous(breaks = seq(2008, 2017, 1), limits = c(2008,2015)) +
  labs(x="Year", y="Incarceration Rate (% of State Population)")
incarc.rate.vis
```


# Table of Incarceration Rates by County and Year
```{r}
county.incarc.rate <- co_vera_data %>% select(year, county_name, total_pop, total_pop_15to64, 
                                             urbanicity, region, total_jail_pop) %>%
  mutate(incarc_perc = total_jail_pop/total_pop_15to64*100)

# Create Table of Incarceration rates by county and year 
county.incarc.table <- county.incarc.rate %>% 
  select(year, county_name, incarc_perc) %>%
  spread(county_name, incarc_perc)

```

# Changes in Incarceration Rate by County
Pay attention to the fact that both changing county population and changing jail population can affect this. 

```{r, fig.height=10, fig.width = 14 }
# Difference in incarceration rates 2015 - 2008
incarc.diff.15.08 <- county.incarc.rate %>% 
  filter(year == 2015 | year == 2008) %>% 
  select(year, county_name, incarc_perc) %>%
  mutate(diff2015.08 = incarc_perc -lag(incarc_perc, default = first(incarc_perc))) %>%
  filter(year == 2015) %>%
  select(county_name, diff2015.08) %>%
  mutate(changesign = ifelse(diff2015.08<0, "Less", "More"))


incarc.diff.15.08.plot  <- ggplot(filter(incarc.diff.15.08, diff2015.08 < 1),
                         aes(x=reorder(county_name, -diff2015.08), y=diff2015.08, fill=changesign)) + 
                        stat_summary(fun.y="sum", geom="bar", position="dodge") + coord_flip() +
                         labs(x="County", y="2015 Incarceration Perc - 2008 Incarceration Perc",
                            title="Change in Incarceration Percentage of County Population (2015 - 2008)"
                                                 ) +   
  theme(
                                                   plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
                                                   axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
                                                   axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
                                                   axis.text.x=element_text(angle=0, size=10, vjust=0.5, family ="serif"),
                                                   axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
                                                   panel.background = element_rect(fill = 'white'),
                                                   panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
                                                   panel.grid.minor = element_line(colour = "white", size = .5)
                                                 ) + 
  guides(fill=FALSE)

incarc.diff.15.08.plot

```



# Comparing Changing Population to Changing Jail Population
This graphs the percent change in the jail population since 2008 against the percent change in the county population since 2008. The line is a reference, not a trend line. Above the line y=x fall counties where incarceration has either increased more than the population or decreased less than the population (thus the incarceration rate has increased). Below the line fall those where incarceration rate has increased less than the population has increased or decreased more than the population has decreased (and thus the overall incarceration rate has fallen)


```{r, fig.height = 14, fig.width = 14}
perc.change.data <- co_vera_data %>%
  select(year, county_name, total_pop_15to64, total_jail_pop) %>%
  filter(year == 2015 | year == 2008) %>%
  mutate(
    dif_15to64_pop = (total_pop_15to64 -lag(total_pop_15to64, default = first(total_pop_15to64)))*100/lag(total_pop_15to64, default = first(total_pop_15to64)),
    dif_jail_pop = (total_jail_pop - lag(total_jail_pop, default = first(total_jail_pop)))*100/lag(total_jail_pop, default = first(total_jail_pop))
    ) %>%
  filter(year == 2015) %>%
  select(county_name, dif_15to64_pop, dif_jail_pop) %>%
  mutate(
    dif_in_dif = dif_jail_pop/dif_15to64_pop,
    changesign = ifelse(dif_in_dif < 0, "Less", "More")
    )


Rel.Change.plot  <- ggplot(filter(perc.change.data, abs(dif_in_dif) < 150 ) ,
                           aes(x= dif_15to64_pop, y=dif_jail_pop, color = as.factor(I(dif_15to64_pop - dif_jail_pop > 0)))) + 
  geom_point()+
  ggtitle('% Change in Incarceration v % Change in Population (2015 - 2008)') +
  theme(
    plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
    axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
    axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
    axis.text.x=element_text(angle=0, size=10, vjust=0.5, family ="serif"),
    axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
    panel.background = element_rect(fill = 'white'),
    panel.grid.major = element_line(colour = "grey", size = .5, linetype = "dashed" ),
    panel.grid.minor = element_line(colour = "grey", size = .3, linetype = "dashed"),
    legend.text = element_text(family = "serif"), 
    legend.title = element_text(family = "serif", face = "bold"),
    legend.position = c(0.2, 0.65)
  ) + 
  ylim(-75, 75) +
  scale_x_continuous(breaks = seq(-20, 20, 5), limits = c(-20,20)) +
  labs(x="% Change in Population", y="% Change in Jail Population", color = "Net Incarceration") +
  geom_text(aes(label=county_name), hjust=-.1, vjust=.1, size = 2) +
  geom_hline(aes(yintercept = 0)) +
  geom_vline(aes(xintercept = 0)) +
  scale_color_hue(labels = c("Increase", "Decrease")) + 
 
 annotate("text", x = c(10,10, -10, -10), y = c(50,-50, -50, 50), label = c("Jail Pop Increase, County Pop Increase","Jail Pop Decrease, County Pop Increase", "Jail Pop Decrease, County Pop Decrease", "Jail Pop Increase, County Pop Decrease"
                                                                            ), family = "serif", fontface = 2)
  





Rel.Change.plot
```







```{r, fig.width = 14}
pretrial_perc_co <- co_jail_pop_by_year %>%
  mutate(pretrial_perc = total_jail_pretrial/total_jail_pop *100) %>%
  select(year, total_jail_pretrial, total_jail_pop, pretrial_perc)

pretrial_perc_plot <- ggplot(pretrial_perc_co, aes(x = year, y = pretrial_perc)) +
  geom_line() + 
  geom_point()+
  ggtitle('Pretrial Jail Population ') +
  theme(
    plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
    axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
    axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
    axis.text.x=element_text(angle=50, size=10, vjust=0.5, family ="serif"),
    axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
    panel.background = element_rect(fill = 'white'),
    panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
    panel.grid.minor = element_line(colour = "white", size = .5)
  ) + 
  ylim(0, 100) +
  scale_x_continuous(breaks = seq(1970, 2015, 1), limits = c(1970,2015)) +
  labs( x="Year", y="% of Jail Population that is Pretrial") 
pretrial_perc_plot

```

``` {r, fig.height = 20, fig.width = 20}
###### County Pretrial Jail Populations 
pretrial_county <- co_vera_data %>% 
  select(year, county_name, total_jail_pretrial, total_jail_pop) %>%
  filter(year > 2012 & year <2016) %>%
  mutate(pretrial_perc = total_jail_pretrial/total_jail_pop *100)

pretrial_county_plot <- ggplot(pretrial_county,
       aes(x=reorder(county_name, -pretrial_perc), y=pretrial_perc, fill=as.factor(year))) + 
  stat_summary(fun.y="sum", geom="bar", position="dodge") + coord_flip() +
  labs(x="County", y="Percent of Jail Population that is Pretrial",
       title="Pretrial Percentage by County and Year" , fill = "Year"
  ) +   
  theme(
    plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
    axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
    axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
    axis.text.x=element_text(angle=0, size=10, vjust=0.5, family ="serif"),
    axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
    panel.background = element_rect(fill = 'white'),
    panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
    panel.grid.minor = element_line(colour = "white", size = .5),
    legend.text = element_text(family = "serif"), 
    legend.title = element_text(family = "serif", face = "bold"),
    legend.position = c(0.9, 0.80)
  ) 

pretrial_county_plot
```
