---
title: "Lincoln County Fact Sheet: Money Bonds"
author: "Charlotte McClintock & Sam Powers"
date: "Fall 2018"
output: 
  html_document:
   toc: true
   theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)

load("moneybail.RData")
co_vera_data <- read_csv("Colorado-Incarceration-Data.csv")
```


## Incarceration Trends in Lincoln County

## Money Bonds vs. PR Bonds in Lincoln County


Lincoln County has set money bond in over 95% of cases for both felonies and misdemeanor every year for the past three years. 


```{r}
ggplot(subset(county, county=="Lincoln"), aes(x=year, y=money_set_perc, fill=off_type)) + 
  stat_summary(fun.y="sum", geom="bar", position = "dodge") + 
  geom_text(aes(x = year, y=money_set_perc, label=round(100*money_set_perc, 1)), position=position_dodge(width = 1), vjust = -0.5) + scale_y_continuous(labels = scales::percent) + annotate("text", x=2014, y=1.02, label=".", color="white") + 
  labs(title="Money Bonds as a % of All Bonds Set in Lincoln County by Year",
       x="Year", y="Money Bonds as a % of All Bonds", fill="Offense Class") + scale_fill_manual(values=c("#003f5c","#bc5090")) 
```

Lincoln County used money bonds more frequently than any other county which processed at least 10 bonds in 2016, for both felony and misdemeanor offenses. 

```{r tab-felony-money-2016}
kable(head(arrange(select(filter(bondset, total_bond>10), 
                          county, year, crime_type, money_perc, total_bond), 
                   -year, crime_type, -money_perc), 10), 
      col.names = c("County", "Year", "Offense Class", "Money Bonds as a % of All Bonds", 
                    "Total Bonds"),
      caption = "Counties With Highest Percentage Money Bond for Felony Offenses in 2016")
```

```{r tab-mis-money}
kable(head(arrange(select(filter(bondset, total_bond>10), county, year, crime_type, money_perc, total_bond), -year, desc(crime_type), -money_perc), 10), col.names = c("County", "Year", "Offense Class", "Money Bonds as a % of All Bonds", "Total Bonds"), caption = "Counties With Highest Percentage Money Bond for Misdemeanor Offenses in 2016")
```

Although many counties decreased their use of money bonds after the passage of HB13.1236, Lincoln County actually modestly *increased* use of money bonds.

```{r, fig.width=8, fig.height=12}
year.plot <- mutate(year.change, 
                      diff=round(diff,3),
                      changesign=ifelse(diff<0, "Less Use of Money Bond", 
                                        "More Use of Money Bond"))
year.plot <- filter(year.plot, (abs(diff)>0 )&bond_type=="Money Bond")


g <- ggplot(year.plot,
            aes(x=reorder(county, -diff), y=diff, fill=changesign)) + 
  coord_flip() +
  stat_summary(fun.y="sum", geom="bar", position="dodge") + 
  facet_wrap(~crime_type) + 
  labs(x="County", y="2016 Money Bail Rate - 2014 Money Bail Rate",
       title="Change in Money Bail as Percentage of All Bonds 2014-2016, by Offense Class",
       fill="Change Direction") + theme(legend.position = "top") + 
  geom_vline(xintercept = 15.5, linetype=3) + 
  geom_vline(xintercept = 16.5, linetype=3) 
g
```









```{r, fig.height=10, fig.width = 14 }

county.incarc.rate <- co_vera_data %>% select(year, county_name, total_pop, total_pop_15to64, 
                                             urbanicity, region, total_jail_pop) %>%
  mutate(incarc_perc = total_jail_pop/total_pop_15to64*100)

# Difference in incarceration rates 2015 - 2008
incarc.diff.15.08 <- county.incarc.rate %>% 
  filter(year == 2015 | year == 2008) %>% 
  select(year, county_name, incarc_perc) %>%
  mutate(diff2015.08 = incarc_perc -lag(incarc_perc, default = first(incarc_perc))) %>%
  filter(year == 2015) %>%
  select(county_name, diff2015.08) %>%
  mutate(changesign = ifelse(diff2015.08<0, "Less", "More"))


incarc.diff.15.08.plot  <- ggplot(filter(incarc.diff.15.08, diff2015.08 < 1),
                         aes(x=reorder(county_name, -diff2015.08), y=diff2015.08, fill=changesign)) + 
                        stat_summary(fun.y="sum", geom="bar", position="dodge") + coord_flip() +
                         labs(x="County", y="2015 Incarceration Perc - 2008 Incarceration Perc",
                            title="Change in Incarceration Percentage of County Population (2015 - 2008)"
                                                 ) +   
  theme(
                      plot.title = element_text(size=15, face="bold", family = "serif", hjust =0.5 ),
                      axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
                      axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
                      axis.text.x=element_text(angle=0, size=10, vjust=0.5, family ="serif"),
                      axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
                      panel.background = element_rect(fill = 'white'),
                      panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
                      panel.grid.minor = element_line(colour = "white", size = .5)
                                                 ) + 
  guides(fill=FALSE) + 
  geom_vline(xintercept = 2.5, linetype=2, color = "red") + 
  geom_vline(xintercept = 3.5, linetype=2, color = "red") 

incarc.diff.15.08.plot

```




