---
title: "Lincoln County Fact Sheet: Money Bonds"
author: "Charlotte McClintock & Sam Powers"
date: "Fall 2018"
output: 
  html_document:
   toc: true
   theme: yeti
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)

load("~/Documents/GitHub/incarceration-trends/Colorado_ACLU/4-money-bail-analysis/moneybail.Rdata")
co_vera_data <- read_csv("Colorado-Incarceration-Data.csv")
```


## Incarceration Trends in Lincoln County

## Money Bonds vs. PR Bonds in Lincoln County


Lincoln County has set money bond in over 95% of cases for both felonies and misdemeanor every year for the past three years. 


```{r}
ggplot(subset(county, county=="Lincoln"), aes(x=year, y=money_set_perc, fill=off_type)) + 
  stat_summary(fun.y="sum", geom="bar", position = "dodge") + 
  geom_text(aes(x = year, y=money_set_perc, label=round(100*money_set_perc, 1)), position=position_dodge(width = 1), vjust = -0.5) + scale_y_continuous(labels = scales::percent) + annotate("text", x=2014, y=1.02, label=".", color="white") + 
  labs(title="Money Bonds as a % of All Bonds Set in Lincoln County by Year",
       x="Year", y="Money Bonds as a % of All Bonds", fill="Offense Class") + scale_fill_manual(values=c("#003f5c","#bc5090")) 
```

Lincoln County used money bonds more frequently than any other county which processed at least 10 bonds in 2016, for both felony and misdemeanor offenses. 

```{r tab-felony-money-2016}
kable(head(arrange(select(filter(bondset, total_bond>10), 
                          county, year, crime_type, money_perc, total_bond), 
                   -year, crime_type, -money_perc), 10), 
      col.names = c("County", "Year", "Offense Class", "Money Bonds as a % of All Bonds", 
                    "Total Bonds"),
      caption = "Counties With Highest Percentage Money Bond for Felony Offenses in 2016")
```

```{r tab-mis-money}
kable(head(arrange(select(filter(bondset, total_bond>10), county, year, crime_type, money_perc, total_bond), -year, desc(crime_type), -money_perc), 10), col.names = c("County", "Year", "Offense Class", "Money Bonds as a % of All Bonds", "Total Bonds"), caption = "Counties With Highest Percentage Money Bond for Misdemeanor Offenses in 2016")
```

Although many counties decreased their use of money bonds after the passage of HB13.1236, Lincoln County actually modestly *increased* use of money bonds.

```{r, fig.width=8, fig.height=12}
year.plot <- mutate(year.change, 
                      diff=round(diff,3),
                      changesign=ifelse(diff<0, "Less Use of Money Bond", 
                                        "More Use of Money Bond"))
year.plot <- filter(year.plot, (abs(diff)>0 )&bond_type=="Money Bond")


g <- ggplot(year.plot,
            aes(x=reorder(county, -diff), y=diff, fill=changesign)) + 
  coord_flip() +
  stat_summary(fun.y="sum", geom="bar", position="dodge") + 
  facet_wrap(~crime_type) + 
  labs(x="County", y="2016 Money Bail Rate - 2014 Money Bail Rate",
       title="Change in Money Bail as Percentage of All Bonds 2014-2016, by Offense Class",
       fill="Change Direction") + theme(legend.position = "top") + 
  geom_vline(xintercept = 15.5, linetype=3) + 
  geom_vline(xintercept = 16.5, linetype=3) 
g
```

## Incarceration Trends in Lincoln County ##
The Jail Population of Lincoln County, as a percentage of the county population, has steadily increased, peaking in 2013. 

```{r, fig.width = 14}
county.incarc.rate <- co_vera_data %>% select(year, county_name, total_pop, total_pop_15to64, 
                                             urbanicity, region, total_jail_pop) %>%
  mutate(incarc_perc = total_jail_pop/total_pop_15to64*100)

incarc.rate.vis <- ggplot(filter(county.incarc.rate, county_name == "Lincoln County"), aes(x= year, y = incarc_perc)) +
  geom_line() + 
  geom_point()+
  ggtitle('Lincoln County Incarceration Rate Over Time') +
  theme(
    plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
    axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
    axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
    axis.text.x=element_text(angle=50, size=10, vjust=0.5, family ="serif"),
    axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
    panel.background = element_rect(fill = 'white'),
    panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
    panel.grid.minor = element_line(colour = "white", size = .5)
  ) + 
  ylim(0, .9) +
  scale_x_continuous(breaks = seq(1970, 2015, 1), limits = c(1970,2015)) +
  labs(x="Year", y="Incarceration Rate (% of State County)")
incarc.rate.vis
```



From 2008 to 2015, incarceration increased by almost 25%. In relationship to other counties, Lincoln county ranks among the greatest increases in incarceration across the state. 

```{r, fig.height=10, fig.width = 14 }



# Difference in incarceration rates 2015 - 2008
incarc.diff.15.08 <- county.incarc.rate %>% 
  filter(year == 2015 | year == 2008) %>% 
  select(year, county_name, incarc_perc) %>%
  mutate(diff2015.08 = incarc_perc -lag(incarc_perc, default = first(incarc_perc))) %>%
  filter(year == 2015) %>%
  select(county_name, diff2015.08) %>%
  mutate(changesign = ifelse(diff2015.08<0, "Less", "More"))


incarc.diff.15.08.plot  <- ggplot(filter(incarc.diff.15.08, diff2015.08 < 1),
                         aes(x=reorder(county_name, -diff2015.08), y=diff2015.08, fill=changesign)) + 
                        stat_summary(fun.y="sum", geom="bar", position="dodge") + coord_flip() +
                         labs(x="County", y="2015 Incarceration Perc - 2008 Incarceration Perc",
                            title="Change in Incarceration Percentage of County Population (2015 - 2008)"
                                                 ) +   
  theme(
                      plot.title = element_text(size=15, face="bold", family = "serif", hjust =0.5 ),
                      axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
                      axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
                      axis.text.x=element_text(angle=0, size=10, vjust=0.5, family ="serif"),
                      axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
                      panel.background = element_rect(fill = 'white'),
                      panel.grid.major = element_line(colour = "grey", size = .3, linetype = "dashed" ),
                      panel.grid.minor = element_line(colour = "white", size = .5)
                                                 ) + 
  guides(fill=FALSE) + 
  geom_vline(xintercept = 2.5, linetype=2, color = "red") + 
  geom_vline(xintercept = 3.5, linetype=2, color = "red") 

incarc.diff.15.08.plot

```


Due to the nature of incarceration as a proportion (Jail Population/County Population), it is affected both by changes in the Jail population and in the County population. This graph breaks all counties into components contributing to incarceration change. 

Lincoln county falls into the second quadrant, which means that while the county population has decreased slightly since 2008, the jail population has increased by about 35%. 

```{r, fig.height = 14, fig.width = 14}


perc.change.data <- co_vera_data %>%
  select(year, county_name, total_pop_15to64, total_jail_pop) %>%
  filter(year == 2015 | year == 2008) %>%
  mutate(
    dif_15to64_pop = (total_pop_15to64 -lag(total_pop_15to64, default = first(total_pop_15to64)))*100/lag(total_pop_15to64, default = first(total_pop_15to64)),
    dif_jail_pop = (total_jail_pop - lag(total_jail_pop, default = first(total_jail_pop)))*100/lag(total_jail_pop, default = first(total_jail_pop))
    ) %>%
  filter(year == 2015) %>%
  select(county_name, dif_15to64_pop, dif_jail_pop) %>%
  mutate(
    dif_in_dif = dif_jail_pop/dif_15to64_pop,
    changesign = ifelse(dif_in_dif < 0, "Less", "More")
    )


Rel.Change.plot  <- ggplot(filter(perc.change.data, abs(dif_in_dif) < 150 ) , 
                           aes(x= dif_15to64_pop, y=dif_jail_pop, color = as.factor(I(dif_15to64_pop - dif_jail_pop > 0)))) + 
  geom_point()+
  ggtitle('% Change in Incarceration v % Change in Population (2015 - 2008)') +
  theme(
    plot.title = element_text(size=15, face="bold", family = "serif", hjust = 0.5 ),
    axis.title.x = element_text(vjust=-0.5, size = 15, family ="serif"),
    axis.title.y = element_text(vjust=0.75,family ="serif", size = 15),
    axis.text.x=element_text(angle=0, size=10, vjust=0.5, family ="serif"),
    axis.text.y=element_text(angle= 0, size=10, vjust=.05,family ="serif"),
    panel.background = element_rect(fill = 'white'),
    panel.grid.major = element_line(colour = "grey", size = .5, linetype = "dashed" ),
    panel.grid.minor = element_line(colour = "grey", size = .3, linetype = "dashed"),
    legend.text = element_text(family = "serif"), 
    legend.title = element_text(family = "serif", face = "bold"),
    legend.position = c(0.2, 0.65)
  ) + 
  ylim(-75, 75) +
  scale_x_continuous(breaks = seq(-20, 20, 5), limits = c(-20,20)) +
  labs(x="% Change in Population", y="% Change in Jail Population", color = "Net Incarceration") +
  geom_text(aes(label=county_name), hjust=-.1, vjust=.1, size = 2) +
  geom_hline(aes(yintercept = 0)) +
  geom_vline(aes(xintercept = 0)) +
  scale_color_hue(labels = c("Increase", "Decrease")) + 
 
 annotate("text", x = c(10,10, -10, -10), y = c(50,-50, -50, 50), label = c("Jail Pop Increase, County Pop Increase","Jail Pop Decrease, County Pop Increase", "Jail Pop Decrease, County Pop Decrease", "Jail Pop Increase, County Pop Decrease"
                                                                            ), family = "serif", fontface = 2) +
  geom_point(data = filter(perc.change.data, county_name == "Lincoln County"), aes(x = dif_15to64_pop, y = dif_jail_pop),  position = "identity", shape = 1, fontface = 4, color = "black", size = 8)


Rel.Change.plot
```